---
title: "Data visualization with ggplot2 (part 1)"
author: "Victoria Restivo"
date: '2018-11-16'
output: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
-NOTE: there are comments in the code as well :) 


#Chapter 1: Introduction to Data Visualization in ggplot2

Data visualization is a form of data analysis, making plots attractive and meaningful

Explorative: easily generated, data heavy, not pretty
Explainatory: intensive, but inteded for a broader audience

Plot/visualize data to show distinct trends

##Load the ggplot package
```{r}
library(ggplot2)
```

If a variable is a category, use factor() - this will help insure values on the axis are only values present in the dataset

NOTE - mtcars is the data set used in data camp
```{r}
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_point()
```

##Grammar of Graphics
Data - dataset being plotted
Aesthetics - scales onto which we map our data
Geometries - visual elements of the data

##make a scatter plot using the data set mtcars, x= the variable weight, y = the variable miles per gallon
```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
geom_point()
```

##Make the color equal to another variable, the displacement (disp) of the car engine
```{r}
ggplot(mtcars, aes(x = wt, y = mpg, color = disp)) +
geom_point()
```
##Make the size of the points equal to disp
```{r}
ggplot(mtcars, aes(x = wt, y = mpg, size = disp)) +
    geom_point()
```

##NOTE - Shape can be used in aes() but doesn't work for continuous variables, only categorical

##Add geom_point() and geom_smooth(), geom_smooth() will draw a smooth line over the points that follows the data

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + geom_point() + geom_smooth()
```


##Assign color = to the variable clarity, must go within the aes()
```{r}
ggplot(diamonds, aes(x = carat, y = price, color=clarity)) +
  geom_smooth()
```


##Plot using geom_points with the argument alpha (alpha makes the points transparent, in this case 40% visible/60% transparent).
##NOTE: color=variable can also be used in the geom_point() function.
##NOTE: you can store a plot as a variable
```{r}
ggplot(diamonds, aes(x = carat, y = price, color=clarity)) +
  geom_point(alpha=0.4)
```

##store plot as variable
```{r}
dia_plot <- ggplot(diamonds, aes(x = carat, y = price))
```

##add geom_point() to the variable above, use aes inside geom_point()
```{r}
dia_plot + geom_point(aes(color = clarity))
```

##use se=FALSE to avoid error shading in geom_smooth()
```{r}
dia_plot + geom_smooth(aes(color=clarity), se = FALSE)
```

#Chapter 2: Objects and Layers

##calculate a linear model using lm()
```{r}
carModel <- lm(mpg ~ wt, data = mtcars)
```

##in base R plot the linear model using abline(), lty() is the line type
```{r}
abline(carModel, lty = 2)
```

##Note: in general ggplot2 is easier to make plots with a legend than base R

```{r}
# Plot 1: add geom_point() to create a scatter plot
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
  geom_point()

# Plot 2: include the lines of the linear models, per cyl
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

# Plot 3: include a lm for the entire dataset in its whole (this will add another trendline for the entire data set rather than the 3 different cyl (categorical variable))
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, linetype = 2) 
```

##You can tell if you are doing something wrong in ggplot if no legend is produced

```{r}
##When layering multiple variables onto one graph, this is preferred (iris.wide is a variable created to include petal length and width, as well as sepal length and width, part=petal or sepal), this will create a proper legend:
ggplot(iris.wide, aes(x = Length, y = Width, col = Part)) +
geom_point()
```

##Tidy Data, easier to plot all different types of species on the same plot so they have the same Y-value, species name can go across the top (i.e. the facet_grid() or facet layer) rather than separate graphs all with different Y-values (which may be hard to compare)

```{r}
str(iris.tidy) 
#gives structure which has all the parts used below (x, y, col, and facet)

ggplot(iris.tidy, aes(x = Species, y = Value, col = Part)) +
    geom_jitter() +
    facet_grid(. ~ Measure) 
#makes a nice plot with a legend, and the title/facet layer is shown allowing for only one y axis and direct comparisons can be made
```
##use functions from tidyr package to convert iris to iris.tidy
##NOTE: data is considered tidy when every row is an observation and every column is a variable

##gather() moves information from columns to rows (also rearrange what is a categorical value using a "-")
##separate() splits one column into two or more
##%>% (pipe) passes the result of the left hand side as the first argument to the function on the right hand side
```{r}
# Load the tidyr package
library(tidyr)

# Fill in the ___ to produce to the correct iris.tidy dataset
iris.tidy <- iris %>%
  gather(key, Value, -Species) %>%
  separate(key, c("Part", "Measure"), "\\.")
```

#Chapter 3: Visible Aesthetics

##call aesthetics in the aes() layer
-x = x axis position
-y = y axis position
-colour = colour of dots/other shapes (also changes outline of an object)
-fill = fill colour (changes filling of an object)


##call attributes in the geom() layer
-size = diameter of points/thickness of lines
-alpha = transparency
-linetype = line dash pattern
-labels = text on a plot or axes
-shape = shape of a point
-size = size of a point

```{r}
# 1 - Map mpg to x and cyl to y
ggplot(mtcars, aes(x=mpg, y=cyl)) +
  geom_point()
  
# 2 - Reverse: Map cyl to x and mpg to y
ggplot(mtcars, aes(x=cyl, y=mpg)) +
  geom_point()

# 3 - Map wt to x, mpg to y and cyl to col, making the points different colours based on the cyl variable (categorical)
ggplot(mtcars, aes(x=wt, y=mpg, col=cyl)) +
  geom_point()

# 4 - Change shape and size of the points in the above plot, changes the size and shape of the points, also made them a clear in the centre of the point
ggplot(mtcars, aes(x=wt, y=mpg, col=cyl)) +
  geom_point(shape=1, size=4)
```

##the default geom_point() uses shape = 19 (a solid circle with an outline the same colour as the inside)
##shape=1 makes the point hollow/no filling (this is what happened in #4 above)
##shape=16 is solid no outline

##shape=21 allows you to use both fill for the inside and col for the outline  **COOL APPLICATION!

```{r}
# From the previous exercise, makes scatter plot with coloured outlines, hollow points
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
  geom_point(shape = 1, size = 4)


# 1 - Map cyl to fill, makes a scatter plot with hollow non-coloured points (i.e. black outline)
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl)) +
  geom_point(shape = 1, size = 4)


# 2 - Change shape and alpha of the points in the above plot, makes scatter plot with colour filled points, 60% visible/40% transparent
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl)) +
  geom_point(shape = 21, size = 4, alpha = 0.6)


# 3 - Map am to col in the above plot, added outlines to the points with a legend (am is a factor variable in the dataset)
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl, col=am)) +
  geom_point(shape = 21, size = 4, alpha = 0.6)
```

```{r}
# Map cyl to size, makes plot with points that change based on size of cyl variable (legend is included)
ggplot (mtcars, aes(x=wt, y=mpg, size=cyl)) + geom_point()

# Map cyl to alpha, points were distinguised with transparency (legend included)
ggplot (mtcars, aes(x=wt, y=mpg, alpha=cyl)) + geom_point()

# Map cyl to shape, different shapes used for cyl variable (legend included)
ggplot (mtcars, aes(x=wt, y=mpg, shape=cyl)) + geom_point() 

# Map cyl to label, different labels for the cyl (either 4, 6, or 8, are layed over the data, no legend)
ggplot (mtcars, aes(x=wt, y=mpg, label=cyl)) + geom_point() + geom_text()
```

##Attributes
-Aesthetics can be used as attributes
-Shapes in R can have a value from 1-25, only shapes 1-20 can accept only a color asthetic, but shapes 21-25 have both a color and a fill asthetic

##Hexadecimal
-Hexadecimal ("related to 16") is a base-16 alphanumeric counting system, hexadecimal colours use a system (00-FF) to specify a six-digit code for red, green and blue values ("#RRGGBB") of a colour, for example: put blue is "0000FF", black "000000", white "FFFFFF")


```{r}
# Define a hexadecimal color
my_color <- "#4ABEFF"

# Draw a scatter plot with color *aesthetic*, makes the typical plot we've been looking at with colours defined by col=cyl, the factor cyl has a legend 
ggplot(mtcars, aes(x=wt, y=mpg, col=cyl)) + geom_point()


# Same, but set color *attribute* in geom layer, the points colour is all not baby blue (the line color=my_color, which was defined in the first line, overwrites col=cyl)
ggplot(mtcars, aes(x=wt, y=mpg, col=cyl)) + geom_point (color=my_color)




# Set the fill aesthetic; color, size and shape attributes, in this one the fill is the 3 different colours we are used to, but the outside of the points are coloured the baby blue "my_colour", the points are large diamonds
ggplot(mtcars, aes(x=wt, y=mpg, fill=cyl)) + geom_point (color=my_color, size=10, shape=23)
```

##NOTE: shape23 = diamonds

```{r}
# Expand to draw points with alpha 0.5 (50% transparent, based on the variable cyl)
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl)) + geom_point(alpha=0.5)

  
# Expand to draw points with shape 24 and color yellow (makes the shape a triangle, and the outline colour yellow, while the cyl variable determines the inside colour with a legend)
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl)) +geom_point (shape=24, colour="yellow")

  
# Expand to draw text with label rownames(mtcars) and color red, writes over the points with the names of the cars in the colour red
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl)) +geom_point () + geom_text(label=rownames(mtcars), color="red")
```

##NOTE: shape=24 is a triangle

##WARNING: Adding more aesthetic mapping to a plot increases its complexity and decreases readibility 

##In general, variables are mapped to aesthetics, visual elements are set by attributes in the geom layer

```{r}
# Map mpg onto x, qsec onto y and factor(cyl) onto col
ggplot(mtcars, aes(x=mpg, y=qsec, color=factor(cyl))) + geom_point()


# Add mapping: factor(am) onto shape
ggplot (mtcars, aes(x=mpg, y=qsec, color=factor(cyl), shape=factor(am))) + geom_point()


# Add mapping: (hp/wt) onto size
ggplot(mtcars, aes(x=mpg, y=qsec, color=factor(cyl), shape=factor(am), size=(hp/wt))) + geom_point()
```

##NOTE: label and shape are only applicable to categorical data, the other ones can be used for categorical or numerical data

##Poistions - identity 
-when identity = default value in the data frame is exactly where we want it to be
-position jitter = can be defined in a function (position_jitter()), or in our ggplot (position = posn.j)

##Scale Functions
-scale_x_continuous OR _discrete (discrete=categorical)
-expand can be used to give a gap between data and axis

```{r}
cyl.am <- ggplot(mtcars, aes(x = factor(cyl), fill = factor(am))) 

# Add geom (position will = "stack" by default, creates a stacked bar-graph of the factors of the variable am with a legend)
cyl.am + geom_bar()

# Fill - show proportion
cyl.am + 
  geom_bar(position = "fill")  

# Dodging - principles of similarity and proximity (place the bar graphs for the factor(am) variable side by side)
cyl.am +
  geom_bar(position = "dodge") 

# Clean up the axes with scale_ functions (legend is more descriptive)
val = c("#E41A1C", "#377EB8") #colours for the bar graphs
lab = c("Manual", "Automatic") #labels for the two different bar graphs 
cyl.am +
  geom_bar(position = "dodge") + #side by side bars
  scale_x_discrete("Cylinders") +  #labels x axis
  scale_y_continuous("Number") + #labels y axis
  scale_fill_manual("Transmission",  #first argument labels the legend
                    values = val, #values are the colours defined above 
                    labels = lab) #labels are the labels above
```

##Setting a dummy aesthetic

```{r}
# 1 - Create jittered plot of mtcars, mpg onto x, 0 onto y (makes univariate plot)
ggplot(mtcars, aes(x = mpg, y = 0)) +
  geom_jitter()

# 2 - Add function to change y axis limits, looks nicer because you changed the y axis limits
ggplot(mtcars, aes(x = mpg, y = 0)) +
  geom_jitter() + scale_y_continuous (limits = c(-2,2))
```

##Important to consider who the reader is and gear plots to them
##Helpful to use alpha blending/transparency to avoid overplotting

```{r}
#Basic scatter plot: wt on x-axis and mpg on y-axis; map cyl to col
ggplot (mtcars, aes(x=wt, y=mpg, col=cyl)) + geom_point(size=4)


#Hollow circles
ggplot (mtcars, aes(x=wt, y=mpg, col=cyl)) + geom_point(size=4, shape=1)

#Transparency 
ggplot (mtcars, aes(x=wt, y=mpg, col=cyl)) + geom_point(size=4, alpha=0.6)
``` 

```{r}
# Scatter plot: carat (x), price (y), clarity (color)
ggplot (diamonds, aes(x=carat, y=price, col=clarity)) + geom_point()


# Adjust for overplotting
ggplot (diamonds, aes(x=carat, y=price, col=clarity)) + geom_point(alpha = 0.5)


# Scatter plot: clarity (x), carat (y), price (color), lines all values along one line in x axis
ggplot (diamonds, aes(x=clarity, y=carat, col=price)) + geom_point(alpha = 0.5)

# Dot plot with jittering, this jitters the values making it easier to read (apparently?)
ggplot (diamonds, aes(x=clarity, y=carat, col=price)) + geom_point(alpha = 0.5, position = "jitter")
```
##jitter is in ""


#Chapter 4: Scatter Plots
-3 common plot types (scatter, bar, line plots)

##Scatter plot
-each geom has specific aesthetic mappings, i.e. geom_point() requires x, y arguments, but there are also optional ones

```{r}
#Use geom_jitter() as a geom:
ggplot(mtcars, aes(x = cyl, y = wt)) +
  geom_jitter()

#Using geom_jitter() and controlling the width:
ggplot(mtcars, aes(x = cyl, y = wt)) +
  geom_jitter(width=0.1)

#Jitter can be an argument in geom_point as a position function
ggplot(mtcars, aes(x = cyl, y = wt)) +
  geom_point(position = position_jitter(0.1))
```

##Histograms - common way of showing distributions

-When using geom_histogram, stat="bin" is the default (histograms cut continuous variables into discrete bins, you always get 30 evenly sized bins specified with binwidth=range/30)

-geom_histogram only requires one aesthetic (x), the y axis comes from ..count.. (it automatically counts how many variables are in each bin - this is stored in an internal data frame

-density is also calculated (proportional frequency, use ..density.. to access)

```{r}
#Univariate (i.e. 1 variable) histogram (y=count)
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram()

#plot above, plus set binwidth to 1 in the geom layer (y=count)
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth=1)


#y=density
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth=1, aes(y=..density..))


#plot above, fill attribute to "#377EB8" (blue)
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth=1, aes(y=..density..), fill="#377EB8")
```

##histograms & bar graphs also have position arguments 
-stack = bars on top of each other, counts used, default
-fill = bars on top, but using proportions
-dodge = bars next to each other, counts used
##NOTE: the positions must be in ""

```{r}
#position = dodge
ggplot(mtcars, aes(x = cyl, fill = am)) +
  geom_bar(position = "dodge")
```


# Definition of a set of blue colors
blues <- brewer.pal(9, "Blues") # from the RColorBrewer package

# 1 - Make a color range using colorRampPalette() and the set of blues
blue_range <- colorRampPalette(blues)

# 2 - Use blue_range to adjust the color of the bars, use scale_fill_manual()
ggplot(Vocab, aes(x = education, fill = vocabulary)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = blue_range(11))
##you can adjust the dodge position so that the bars overlap using position_dodge(width=___)

```{r}
#Define posn_d with position_dodge()
posn_d <- position_dodge(width=0.2)

#Use the position argument to posn_d (bars overlap)
ggplot(mtcars, aes(x = cyl, fill = am)) +
  geom_bar(position = posn_d)


#Use posn_d as position and adjust alpha to 0.6 (cool overlap and transparency)
ggplot(mtcars, aes(x = cyl, fill = am)) +
  geom_bar(position = posn_d, alpha=0.6)
```


##You can also overlap a histogram, but you need to turn it into a frequency polygram

```{r}
# A basic histogram, add coloring defined by cyl
ggplot(mtcars, aes(mpg, fill=cyl)) +
  geom_histogram(binwidth = 1)

# Change position to identity, this is messed up though
ggplot(mtcars, aes(mpg, fill=cyl)) +
  geom_histogram(binwidth = 1, position="identity")


# Change geom to freqpoly (position is identity by default), this shows the histogram/new frequency polgram into something neat
ggplot(mtcars, aes(mpg, color=cyl)) +
  geom_freqpoly(binwidth = 1, position="identity")
```

```{r}
##in bar graphs you can set the colour using "scale_fill_brewer"

# Definition of a set of blue colors
blues <- brewer.pal(9, "Blues") # from the RColorBrewer package

# 1 - Make a color range using colorRampPalette() and the set of blues
blue_range <- colorRampPalette(blues)

# 2 - Use blue_range to adjust the color of the bars, use scale_fill_manual()
ggplot(Vocab, aes(x = education, fill = vocabulary)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = blue_range(11))
  
```

##geom_line() plots
-Used for line series
-Follow same syntax as geom_point/geom_bar/geom_histogram
-Series can be encoded with size, line type, or colour

##used in geom_line, geom_area() default is stack, basically fills in the lines under neath, if you use "fill" it does proportion (you would need to change the y axis since its no longer counts!!)

##geom_ribbon helps to show uncertainty, and allows us to show overlapping series (data for variables)

##geom_rect needs 4 aes's: xmin, xmax, ymin, ymax

```{r}
# Basic line plot
ggplot(economics, aes(x = date, y = unemploy/pop)) +
  geom_line()

# Expand the following command with geom_rect() to draw the recess periods
ggplot(economics, aes(x = date, y = unemploy/pop)) +
  geom_rect(data = recess, 
         aes(xmin = begin, xmax = end, ymin=-Inf, ymax=+Inf),
         inherit.aes = FALSE, fill = "red", alpha = 0.2) +
  geom_line()
```

#Chapter 5: qplot
-quick and easy way to generate plots
-seems like less variables to specify in code
-qplot picks the geom for us
-you can still use jitter to correct for overplotting
-alpha plotting doesn't quite work because it adds to the legend, you can get around this by doing: alpha=I(0.5)

```{r}
# Using ggplot:
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point()

# Using qplot:
qplot(x=wt, y=mpg, data = mtcars)

##both of the above plots give the same output
```

```{r}
# basic qplot scatter plot:
qplot(wt, mpg, data = mtcars)

# Categorical variable mapped onto size:
# cyl
qplot(wt, mpg, data = mtcars, size = factor(cyl))

# gear
qplot(wt, mpg, data = mtcars, size = gear)

# Continuous variable mapped onto col:
# hp
qplot(wt, mpg, data = mtcars, color = hp)

# qsec
qplot(wt, mpg, data = mtcars, color = qsec)
```

```{r}
# qplot() with x only
qplot(factor(cyl), data = mtcars)

# qplot() with x and y
qplot(factor(cyl), factor(vs), data = mtcars)

# qplot() with geom set to jitter manually
qplot(factor(cyl), factor(vs), data = mtcars, geom = "jitter")
```


```{r}

# "Basic" dot plot, with geom_point():
ggplot(mtcars, aes(cyl, wt, col = am)) +
  geom_point(position = position_jitter(0.2, 0))

# 1 - "True" dot plot, with geom_dotplot():
ggplot(mtcars, aes(cyl, wt, fill = am)) +
  geom_dotplot(binaxis = "y", stackdir = "center")

# 2 - qplot with geom "dotplot", binaxis = "y" and stackdir = "center", this is the same as the ggplot above
qplot(
  cyl, wt,
  data = mtcars,
  fill = am,
  geom = "dotplot",
  binaxis = "y",
  stackdir = "center"
```
  
  
```{r}
#ggplot
ggplot(titanic, aes(x = Pclass, fill = Sex)) +
  geom_bar(position = "dodge")

#add facet_grid() layer (makes the two different graphs side by side with one y axis, and titles to differentiate the graphs)
ggplot(titanic, aes(x = Pclass, fill = Sex)) +
  geom_bar(position = "dodge") + facet_grid(. ~ Survived)

#Define an object for position jitterdodge, to use below
posn.jd <- position_jitterdodge(0.5, 0, 0.6)

# Use the position object above (overlays points (in above posn.jd definition), transparency 50%))
ggplot(titanic, aes(x = Pclass, y=Age, color = Sex)) +
  geom_point(size=3, alpha=0.5, position = posn.jd) + facet_grid(. ~ Survived) 
```


